Directionality and Direction of Innovation in the Swedish Forest-Based Bioeconomy Transition

### 2024-03-13 paper.qmd
migration commit

---
title: Shades of Green in the Swedish Forest-Based Bioeconomy Transition 
subtitle: Draft
author: 
    name: Jonas Kreutzer
    affiliation: Department of Economic History, Lund University
date: last-modified
abstract: |
    The transition to a bioeconomy is high on the agenda for research, industry and policy. Beyond conceptual uncertainties of what is meant by a bioeconomy, three different visions have emerged, differing in their aims and objectives, drivers and mediators of innovation and relation to the biosphere. The use of detailed innovation data from a Swedish innovation database, SWINNO, allows to capture the actual innovation output of the Swedish forest based bioeconomy innovation system. Qualitative coding of the underlying source material associates innovation from 1970 to 2010 with visions of a bioresource, biotechnology and / or bioecology bioeconomy. Preliminary results suggest that the innovation output of the forest based bioeconomy has decreased relative to the total output. Of the three visions, the bioresource vision has been the most dominant, especially in the 1970s and after 2005. At the same time, innovation which fit into multiple visions exist, suggesting that the opportunity space for future development is not yet locked in.

csl: https://zotero.org/styles/apa?source=1
bibliography: [../../swinno-shades-green.bib]
link-citations: true
link-bibliography: true
colorlinks: true
number-sections: true
format:
  hikmah-manuscript-pdf: 
    papersize: A4
execute:
    include: true
    echo: false
    warning: false
---

@@

The Baltic forests are facing a trilemma of expectations to mitigate climate change, preserve biodiversity, render other ecosystem services and provide the resources for a future bioeconomy [@hogbom2021TrilemmaNordicBaltic].
Balancing these expectations includes managing conflicting goals and hence difficult trade-offs between ecosystem services. 
Transitioning to a bioeconomy, meaning an economy based on bio-based rather than fossil-based inputs is set to increase the pressure on Baltic forests in the coming years [@hogbom2021TrilemmaNordicBaltic].
In recent years transitioning to a bioeconomy has gained relevance among policymakers, practitioners and researchers. 
Connected to this are hopes of increased resource efficiency and renewability.
In a report commissioned by the Swedish Government, the @2012StrategyBiobasedEconomy defines a vision of a Swedish bioeconomy based on two central pillars.
It is to be an economy based on,

@@

These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread and dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic]. Despite -- or perhaps because of -- bioeconomy lacking a strong definition, the term is used to refer to different concepts, with different aims and objectives, foci, normative assumptions and hence implications for society and environment [@bugge2016WhatBioeconomyReview].

@@

The exact role played by forests in a bioeconomy transition depends on the desired vision [@kleinschmit2014ShadesGreenSocial]. There is no doubt that the Swedish vision places high demands on forests as primary providers of the required biomass. Both in the strategy formulated in @research2012StrategyBiobasedEconomy and in public discourse, a small, well-connected group of actors center the bio-resource vision as legitimate and desirable [@holmgren2022WhoseTransformationThis]. The result is a forest sector which operates under a "more of everything" paradigm [@belandlindahl2015SwedishForestryModel].

## Innovation for a Bioeconomy {#sec-innovation_bioeconomy}

Apart from the substitution of fossil resources with bio-based resources, a central narrative used in the Swedish transition is that, "\[t\]echnological innovation is key to a greener future" [@holmgren2022WhoseTransformationThis 42]. According to @jankovsky2021InnovationsForestBioeconomy there are four types of innovation crucial for a forest-based bioeconomy:

1.  Substitute Products,
2.  New (bio-based) Processes,
3.  New (bio-based) Products,
4.  New Behavior.

Currently, most research into bioeconomy innovation relies on patent data or qualitative case studies [e.g., @bennich2021BiobasedEconomy2030; @jander2019MonitoringTransitionBioeconomy]. Unfortunately, these approaches are not without their flaws. Concentrating on patents might overemphasize the significance of the biotechnology vision, which predominantly relies on a science-oriented innovation approach driven by patents [@bugge2016WhatBioeconomyReview]. On the other hand, qualitative case studies, may miss broader trends and features of a bioeconomy transition.

Public, scientific and political debate of bioeconomies has brought forth a number of discourse analyses.
Both @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy conclude that the concept of a bioeconomy conceals different, possibly conflicting visions of transformation outcomes.
Recent scholarship has called for advancing research fronts instead of remaining entrenched in contested discourse positions [@allain2022BioeconomicsVsBioeconomy].

This paper takes up this call and asks what bioeconomy the Swedish innovation system pursues. A unique database of actual innovation output is first used to answer how innovative the system has been between 1970 and 2020. The database sources are then used to provide a qualitative insight into the extent to which specific bioeconomy visions appear in commercialized innovation and the extent to which they include eco-innovations which could lessen environmental impact.

@@

Innovations were classified as belonging to a bioeconomy vision if they fulfilled one, or more criteria outlined in @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy.
@tbl-bugge and @tbl-vivien show the summary of bioeconomy visions presented in the cited papers.
For the remainder of the paper the nomenclature of @bugge2016WhatBioeconomyReview is adopted.
Vivien et al.'s [-@vivien2019HijackingBioeconomy] Type I bioeconomy is taken to correspond with the bioecology vision, Type II and Type III to correspond with the bioresource and biotechnology vision, respectively.
The analysis of bioeconomy discourse which revealed different visions, highlights that the visions have emerged at different times and with different actors at their centers [@vivien2019HijackingBioeconomy, @bugge2016WhatBioeconomyReview].
@vivien2019HijackingBioeconomy trace the introduction of the term bioeconomics to the 1920s and a work on fishery by the Russian biologist Baranoff.
The term bioeoconomics which they label as Type I bioeconomy, was introduced by Georgescu-Roegen, in [-@georgescu-roegen1975BioeconomicAspectsEntropy].
@vivien2019HijackingBioeconomy argue that the term biotechnology gradually emerged between 1990 and 2000, in the wake of enthusiasm about biotechnological advances, such as the discovery of DNA.
It gained popularity and international importance when the OECD began increasingly supporting and reshaping its meaning at the turn of the millennium [@vivien2019HijackingBioeconomy].
According to them, the last concept to enter the discourse was the bioresource oriented bioeconomy.
While biomass has been used throughout history in various ways, the renewed interest with the intention to replace fossil inputs, for example in European Commission communication since 2010.
However, @vivien2019HijackingBioeconomy point to the long history of biorefinery as a core aspect of this vision, which was discussed as early as the 1930s.
Hence, while the organization of the discourse into the labels biotechnology, bioecology and bioresource vision is a very recent research outcome, the underlying concepts have a richer history.
This alleviates concerns that using these labels retroactively confines the innovations to current policy terms.

@@

@tbl-vision_examples shows five examples of classified innovation.
This selection also highlights that multiple labels can be attached to one innovation.
The last innovation, describing a waterproof paper, is classified as a bioresource vision innovation because it describes a process by which biomass is converted and upgraded.
It is also classified as a biotechnology innovation since, according to the innovation source article, the process mimics trees' microfiber formation.
Lastly, there is also the possibility that an innovation cannot be placed into at least one of the visions.
This is the case for innovation which brings general technological improvements, purely advances economic performance, or relates to a general purpose technology such as appliance of computers.
These innovations are labeled as No Bioeconomy Vision, which does not imply that the innovation is not used, or originates in the forest based bioeconomy as defined above.
An illustrative example of such an innovation are improved work boots, which promote safety of forest workers but do not correspond with any of the criteria in @tbl-bugge or @tbl-vivien.

```{python}
# | label: tbl-vision_examples
# | tbl-cap: "List of Eco-Innovation Labels"

@@

This initial list was tested on a subset of innovation in SWINNO and augmented with 15 categories.
Because the list of eco-innovation types was relatively exhaustive from the beginning, few adjustments had to be made.
These include: expanding the labeling for other sustainability criteria, such as human health and safety, and specific forest related aspects / tasks to gain a more detailed understanding of an innovations aim.
For example, some innovations aim to protect harvested timber, either at harvest, during transport or storage.
While @tbl-full_codes shows all new added labels, in the following text those that do not clearly represent an eco-innovation -- following @garcia-granero2018EcoinnovationMeasurementReview "…fewer adverse effects on the environment and more efficient use of resources" (p. 305) -- are excluded.
These excluded labels are Gen-manipulated plant, Increased plant yield, Gen-manipulated animal (incl new breed), Human Health and Safety, Automatization, Precision, and Biocide.

@@

```{python}
# | label: fig-bioeconomy_share
# | fig-cap: "Bioeconomy Innovation Trends in SWINNO (Uncleaned)"
# | fig-subcap:
# |   - "Bioeconomy Innovation Count"
# |   - "Bioeconomy Innovation Share in %"
# | layout-ncol: 1

@@

bioeconomy_visions.loc[
    bioeconomy_visions["bioeconomy_vision"] == "Not Bioeconomy",
    ["bioeconomy_vision"],
] = "No Bioeconomy Vision"

@@

```{python}
# | label: fig-bioeconomy_trends
# | fig-cap: "Count of Bioeconomy Visions by Year"

@@

@fig-bioeconomy_trends illustrates that the peaks in innovation output, identified in @fig-bioeconomy_share and discussed above, were driven by innovation related to the Bioresource Vision as well as No Bioeconomy Vision.
While the Bioresource Visions did strongly contribute to the first peak, the second appears to have been primarily driven by innovation which cannot be associated as belonging to any vision of a bioeconomy.
This raises a couple of questions regarding the effectiveness of past and current policy.
It does add, for example, additional heft to the argument put forth by @holmgren2022WhoseTransformationThis that industry interest dominate the transformation to a bioeconomy.
Further, the persistent dominance of the bioresource vision over the full study period suggests that there might not be much effort, at least in terms of production of innovations, put into creating a transformation towards a bioeconomy.

```{python}
# | label: fig-vision_heatmap
# | fig-caption: "Sector's Contribution to Bioeconomy Visions"
# | include: false

@@

# sectors.loc[sectors["vision"] == "Not Bioeconomy", "vision"] = "No Bioeconomy Vision"

>>>>

---
title: Directionality and Direction of Innovation in the Swedish Forest-Based Bioeconomy Transition 
subtitle: Draft
author: 
    name: Jonas Kreutzer
    affiliation: Department of Economic History, Lund University
date: last-modified
abstract: |
    The transition to a bioeconomy is high on the agenda for research, industry and policy. Beyond conceptual uncertainties of what is meant by a bioeconomy, three different visions have emerged, differing in their aims and objectives, drivers and mediators of innovation and relation to the biosphere. The use of detailed innovation data from a Swedish innovation database, SWINNO, allows to capture the actual innovation output of the Swedish forest based bioeconomy innovation system. Qualitative coding of the underlying source material associates innovation from 1970 to 2010 with visions of a bioresource, biotechnology and / or bioecology bioeconomy. Preliminary results suggest that the innovation output of the forest based bioeconomy has decreased relative to the total output. Of the three visions, the bioresource vision has been the most dominant, especially in the 1970s and after 2005. At the same time, innovation which fit into multiple visions exist, suggesting that the opportunity space for future development is not yet locked in.

csl: https://zotero.org/styles/apa?source=1
bibliography: [swinno-shades-green.bib]
link-citations: true
link-bibliography: true
colorlinks: true
number-sections: true

execute: 
  warning: false
  error: false
  echo: false

format:
    pdf:
        papersize: A4
        linestretch: 1.25
---

@@

<!-- Paragraphs 1-2: Motivation. After reading these paragraphs a reader in any field of economics should believe that if you answer your research question your paper will make an important contribution. -->
Discussions about transitions to a bioeconomy are a hot topic of research and policy at the moment.
Different visions of the endgoal exists, stressing needs for ecological, resource or technological aims.
Actors in the transition stress the need for innovation to transition to a bioeconomy.
Yet, the actual efforts undertaken to realize these visions remain underexplored.

<!-- • Paragraphs 3-4: Challenges. These paragraphs explain why your research question has not already been answered, i.e., what are the central challenges a researcher must tackle to answer this question. -->
Rigorously operationalizing directionality is challening due to a plurality of possible development paths [@stirling2011PluralisingProgressIntegrative].
Stirling argues that innovation is better understood as a vector of possibilities rather than a scalar decision between clearly defined alternatives.
<!-- Lastly, Stirling (2011, 2009, 2008) is a clear voice pointing out a slightly different implication of directionality. Arguably, he introduced the concept to the field of sustainability transitions research, as part of broader critique of one-dimensional growth ideology. In line with a multidimensional view on sociotechnical change, he stresses that innovation is better viewed as “a vector, rather than just a scalarquantity” (Stirling, 2011, p. 84, 2008, p. 263). But he goes further and makes a passionate and eloquent call for a more pluralistic perspective on technological innovation.

@andersson2021OutcomesDirectionalityMorphology 111 -->
Where innovation indicators are concerned, directionality is frequently constrained to a dichotomous choice by the innovator, for example, to produce an environmental innovation or to produce a conventional innovation.
In reality, however, innovators face a much more uncertain decision with many possible paths, of which some might be exclusionary.
Additionally, these studies frequently focus on patents as proxies for innovation acitvity.
Patents, however, suffer from the flaw that they may be filed for strategic reasons for example, to prevent a competitor from accessing a technolgy.
Care should therefore be taken to infer the direction of a transition from patent fillings alone.

Previous research from transition studies and political sciences has been more succesful in illuminating the pluarlity of possible tranisiton directions.
However, the basis for this is usually confined to the analysis of discourse, which can illuminate the intent, but as the saying goes "Talk is cheap".
Without investigating what actors are willing to implement discourse analysis remains a poor measure of directionality in innovation directionality.

A central challenge in assessing the directionality of transitions lies in lack of high quality data that allows to unpack which of the intended directions are seriously pursued by innovators.

<!-- • Paragraph 5: This Paper. This paragraph states in a nutshell what the paper accomplishes and how. -->
This paper adresses this issue by using a unique database of significant Swedish innnovations to examine trends in the transition to a forest based bioeconomy.
Technical source documents are qualitatively coded regarding their alignment with different visions of a bioeconomy.
The result paints a detailed picture of activities to achieve competing visions and through narrative analysis highlights how development blocks enable and prevent certain transition directions.

<!-- • Paragraphs 6-7: Model. Summarize the key formal assumptions you will maintain in your analysis. -->
The core assumption needed for this paper is that innovations can be aligned with visions of transitions without requiring that the vision is already explicitly formalized.

<!-- • Paragraphs 8-9: Data. Explain where you obtain your data and how you measure the concepts that are central to your study. -->
Bioeconomy innovations are identified through SNI codes of user and producer sectors as well as keywords in innovation descriptions from the SWINNO database.
SWINNO consists of technical newspaper articles which report on innovations for a professional audience and hence offer high quality data on the type, function, construction and aims of or motivation behind innovations.
Because these articles are published in edited professional journals they represent innovations of significance for the market and hence accurately capture the direction of search deemed most promising by the innovation system.

<!-- • Paragraphs 10-11: Methods. Explain how you take your model to the data and how you overcome the challenges you raised in paragraphs 3-4. -->
The inclusion of an innovation is based on the value chain of the forest based bioeconomy [@wolfslehner2016ForestBioeconomyNew].
I develop a classification scheme build on visions of bioeconomy transition outcomes and eco-innovation types and apply it to the database sources.
This allows me to analyze innovation as a vector rather than a scalar as called for by @stirling2010MulticriteriaDiversityAnalysis.

<!-- • Paragraphs 12-13: Findings. Describe the key findings. Make sure they connect clearly to the motiva-
tion in paragraphs 1-2. -->
I find that the Swedish bioeconomy innovation output has declined over the last 50 years.
Of the three different visions the bioresource vision is the most pronounced, in line with expecatations from discourse analysis and the Swedish bioeconomy strategy.
I show, however, also that the strategy should not be seen as visionary, as the innovation system had been producing bioeconomy innovation and even bioresource aligned innovation long before the strategy was formulated.
This points to a strong interconnection between path dependence and the creation of directionality in the policy space.
Suprisingly, I find that the majority of innovations related to the bioeconomy cannot be classified as aligning with one of the three visions.
Instead they represent general improvements to machinery and tools used in various segments of the forest based bioeconomy.

The bioresource vision is rooted in a development block to respond to landscape pressure of energy scarcity during the 1970s oil crises.
Examples of a strong bioecological vision of high value added products with regional identity exists, but lack coordination and remain in the niche.

<!-- • Paragraphs 14-15: Literature. Lay out the two main ways your paper contributes to the literature. Each paragraph should center around one contribution and should explain precisely how your paper differs from the most closely related recent work. -->
This paper adds to the literature on directionality by providing a deep insight into the direction of search for an ongoing sustainability transition.
<!-- From an empirical perspective, this is essentially about furthering the current understanding of how sociotechnical systems grow as a result of historical path dependencies, system-internal dynamics and external influence, towards a more elaborate understanding of how these mechanisms influence how system configurations change over time. From a conceptual perspective, it is rather about enabling such investigations, by developing analytical frameworks that capture not only the pace, but also the directionality of innovation processes. For example, there may be a need to advancethe commonly used technological innovation systems framework, since the typology of functional processes is geared towards investigating system growth in a way that hides the potential for vastly different technical, social and spatial development trajectories (Andersson, 2020).
@andersson2021OutcomesDirectionalityMorphology [127-128] -->

It also adds to the literature about bioeconomy transitions specifically by measuring progress towards different visions of a bioeconomy for the first time using innovations produced by the Swedish forest based bioeconomy.

# Directionality, Direction and the Transition to a Bioeconomy {#sec-literature}

This section reviews the literature on transitions to a bioeconomy, introduces competing visions of desired outcomes of this transition, central actors in the Swedish discourse and key concepts relating to the governance of sustainability transitions such as the transition to a bioeconomy, namely directionality and the direction of search within innovation systems.

## Clarification of Bioeconomy Concept

The bioeconomy has gained relevance among policymakers, practitioners and researchers alike.
Broadly speaking, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources.
<!-- TODO: citation -->
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficent wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other eco-system services such as forest biodiversity.
Resolving these tradeoffs is challening, as goal conflicts between different development pathways make simulatneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

As one of the leading forestry nations in the world, Sweden sees substantial economic and political potential in advancing the bioeconomy .
In a report commissioned by the Swedish Government, the @2012StrategyBiobasedEconomy lays out the plan for a Swedish bioeconomy based on two central pillars.
It is to be an economy based on,

@@

These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread and dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic].

### Bioeconomic Visions {#sec-bioeconomy-visions}

@@

An overview or more likely more elaborate description of @vivien2019HijackingBioeconomy and @bugge2016WhatBioeconomyReview here.
Tables as placeholders for now.

The exact role played by forests in a bioeconomy transition depends on the desired vision [@kleinschmit2014ShadesGreenSocial]. There is no doubt that the Swedish vision places high demands on forests as primary providers of the required biomass. Or, as @belandlindahl2015SwedishForestryModel call it a forest sector that operates under a "more of everything" paradigm. Both in the strategy formulated in @research2012StrategyBiobasedEconomy and in public discourse, a small, well-connected group of actors center the bio-resource vision as legitimate and desirable [@holmgren2022WhoseTransformationThis].

A paragraph here explaining the detailed findings of @holmgren2022WhoseTransformationThis discourse network here.

Despite -- or perhaps because of -- bioeconomy lacking a strong definition, the term is used to refer to different concepts, with different aims and objectives, foci, normative assumptions and hence implications for society and environment [@bugge2016WhatBioeconomyReview].
Common ground in ambitions to realize a bioeconomy is the need for diverse types of social, technological and organizational innovation [@el-chichakli2016PolicyFiveCornerstones].  According to @jankovsky2021InnovationsForestBioeconomy there are four types of innovation crucial for a forest-based bioeconomy: (i) substitute products, (ii) new (bio-based) processes, (iii) new (bio-based) products, (iv) new behavior.
The need for innovation also features large in Swedish bioeconomy discourse and narratives therein [@bauer2017TechnologicalInnovationSystems, @bennich2021BiobasedEconomy2030], as, "\[t\]echnological innovation is key to a greener future" [@holmgren2022WhoseTransformationThis 42].

## Visions Provide Directionality to Innovation for Transitions

Visions, narratives, mental models and expectations about technological innovation play a central role in social science sustainability research such as the literature on innovation systems, transition studies or economic policy models such as Mission Oriented Innovation Policy.

<!-- ### Plurality of Possible Development Pathways
Not infinite, as transition already embedded in time and space [@andersson2021OutcomesDirectionalityMorphology], but more than scalar @stirling2010MulticriteriaDiversityAnalysis, @stirling2011PluralisingProgressIntegrative, @bergek2015TechnologicalInnovationSystems -->

### Directionality

As innovation policy took a normative turn under which innovation is needed to solve grand societal challenges [@schot2018ThreeFramesInnovation], the question of how to steer innovation into desired directions became a central issue in innovation research [@kohler2019AgendaSustainabilityTransitions].
Innovation turned from a scalar concept, where the central questions where how to promote the right amount of innovation, into a question of influencing the vector in which society develops through innovation [@stirling2011PluralisingProgressIntegrative].
But, as innovation systems are embedded in a spatial and temporal context they tend to develop along trajectories characterized by path dependency [@kemp2022DiversifyingDeepTransitions, @bergek2015TechnologicalInnovationSystems].

Two concepts are central to understanding how innovation is supposed to be guided based on normative decisions between different feasible paths, both serve as the background concepts in this and many related studies on sustainability transitions.
The multilevel perspective (MLP) introduced by [@geels2007TypologySociotechnicalTransition, @geels2010OntologiesSociotechnicalTransitions] and the literature on innovation systems at their diverse levels of analysis such as technological, regional, or national levels.
<!-- TODO add citations. -->
@hekkert2007FunctionsInnovationSystems provide a framework of intervention in innovation on a systemic level, which highlights the importance of guidance of search as one of seven functions innovation system must provide.
They explain guidance of search as, "those activities within the innovation system that can positively affect the visibility and clarity of specific wants among technology users" (423).
It is changing preferences in society, which through market and government influence and, ".. an interactive and cumulative process of exchanging ideas between technology producers, technology users, and many other actors" (423) decides which direction of technological development is most promising. 
@bauer2017TechnologicalInnovationSystems adapts the innovation system functions to the specific case of biorefineries in Sweden and define the influence on the direction of search as "the incentives and/or pressures for organizations to enter the technological field. These may come in the form of visions, expectations of growth potential, regulation, policy targets, standards, articulation of demand from leading customers, crises in current business, etc." (4).

<!-- 
The most accepted view in sustainability transitions studies is that system change is driven by a combination of exogenous and endogenous driving forces. Moreover, the external shocks ‘do not mechanically impact niches and regimes, but need to be perceived and translated by actors to exert influence’ (Geels & Schot, 2007, p. 404). This implies that the actual directions of change are shaped by strategic action (Grillitsch et al., 2018; Markard & Truffer, 2008; Pacheco et al., 2010; Smith et al., 2010; Yap & Truffer, 2019). Although there are many studies on the role of agency in shaping sustainability transitions, there is still limited understanding about how this agency shapes the directionality of the change process.
@yang2022ShapingDirectionalitySustainability [752]
-->

<!-- 
The directionality battle is not about whether the new (niche actors) will win over the old (regime actors). In our research we do not want to tie regime actors upfront to a strategy of maintaining institutions (defending the regime) while niche actors do the creating (building niches) and disrupting work (destabilizing regimes). Rather, battles about the actual course of action may happen equally among regime actors within a prevailing regime as among actors supporting (potentially manifold) niches. Such a view accounts for a situation in which regime actors may operate in the niche and have an interest in promoting niches, while niche actors may not want to destroy the regime and prefer to operate at the niche level only. In the end, the unfolding directionality will be the result of interaction among actors with different degrees of incumbency (Jørgensen, 2012; Yap & Truffer, 2019). The fact that regime actors are not just defending the status quo has also been recognized in neo-institutional literature. The seminal work of Leblebici et al. (1991) emphasized that internal institutional contradictions may emerge as a starting point for dominant actors to engage with institutional change. In transition studies, Fünfschilling and Truffer (2014) elaborated how different institutional logics in a regime may create tensions within and among actors who are incumbents in the prevailing regime. We have therefore to account for a multiplicity of institutional work strategies of a multitude of actors, which are more or less tied to the prevailing regime structures.

@yang2022ShapingDirectionalitySustainability [754] -->

<!-- Our analysis supports that in the context of transformative industrial change, companies’ innovation activities change directionality. Companies’ central roles in the RIS still lie in exploiting, combining, and recombining knowledge and turning it into innovations (Autio, 1998; Isaksen & Trippl, 2016). The influence of the direction of search and legitimacy of change provided by pioneering organizations (Grillitsch & Sotarauta, 2020), such as Paper Province, additionally supports their identification with the environmental debate.

@martin2023ReconsideringActorRoles 1643 -->

@yang2022ShapingDirectionalitySustainability too highlight the importance of visions in directionality over transformations, stressing the importance of shared visions between niche and regime actors.
They argue that a stretch-and-transform pattern is more likely if niche actors are given a leading role and core regime actors obtain new roles and identites.
"The objective in the stretching and transforming form of niche empowerment is to convince the wider world that the rules of the game need to be changed.", @smith2012WhatProtectiveSpace [17].

Visions as shared imaginaries, mental models and aspirations,
narratives as their embodiment in discourse [@bauer2018NarrativesBiorefineryInnovation, @holmgren2022WhoseTransformationThis], technology as their material realization.

@konnola2021TransformativeGovernanceInnovation highlight the importance of shared visions to provide a normative direction and purpose to a system.

@grillitsch2019InnovationPolicySystemwidea illustrate that visions alone are not sufficent to provide clear directionality.
They must be specific enough and with significant buyin from actors, which may be difficult to obtain with vision statements created elsewhere than within the system.

<!-- Especially, this study indicates that the build-up of shared visions across niche and regime actors is key, and when these shared visions allow for a leading role of niche actors combined with openings for new roles and identities of core regime actors, the emergence of a stretch-and-transform pattern is more likely.

@yang2022ShapingDirectionalitySustainability [766] -->

<!-- Key paper cited in every paper on directionoality
@weber2012LegitimizingResearchTechnology -->

Instead of linear view, all levels interact to provide directionality, i.e. influencing the direction of search.
Other influences on the direction of search exists as well, such as landscape pressures, positive and negative pressures, windows of opportunties etc.
Top down visions and bottom up pragmatism interact to shape transition dynamics [@bolton2019GrandVisionsPragmatic].
Operationalizing directionality hence difficult.
Few notable attempts such as
@cappellano2022WhereInnovationMeets operationalize directionality as a measure of R&I expenditure earmarked for specific transition objectives.
Other attempts to measure the directionality and progress towards a bioeconomy @jander2019MonitoringTransitionBioeconomy and @jander2020MonitoringBioeconomyTransitions, but lack the plurality of innovation.

<!-- As Stirling (2011) emphasizes, transition pathways are contingent. They involve contestations between different worldviews and interests. Instead of focusing entirely on the cascades of innovation that converge into clear-cut and singular regime shifts, transition analysis should remain attentive to the multiplicity of directions that a transition process may take. To be sure, the DT frameworkworks with an evolutionary logic that is far from deterministic. It addresses directionality explicitly. Schot & Kanger (2018, p. 8) consider how the diversity of current socio-technical niches make for undercurrents that contain the seeds for alternative, sustainable and just directions of development. Kanger & Schot (2018, pp. 15-16) describe a ‘portfolio of directionality’ that, throughout subsequent surges and notwithstanding the alignments across sectors, becomes broader over time. Along the formation of ‘regimes’ and ‘meta-regimes’, niche alternatives are not considered to just perish or become absorbed. Instead, these alternative directions may manifest after periods of hibernation, or they break through locally.
@kemp2022DiversifyingDeepTransitions [112] -->

<!-- However, the DT framework displays a rather reductionist understanding of directionality. It reduces the matter to analytical openness, minding the evolutionary multiplicity of a transformation trajectory (e.g. towards, or away from, mass production). The normative reflection remains tied to a particular direction that a transition process should be taking (K ̈ ohler et al., 2019). Yet what is needed is a more fundamentalattending to the multiplicityof possible socio-technical configurations that may emerge,to the different process dynamics that may unfold, and to the different (competing) normative appraisals – by stakeholders and by analysts - of outcomes (Pel et al., 2020). Stirling (2009, 2011) underlined the need to unfold evolutionary diversity in a way that clarifies the normative implications of this diversity.To articulate directionality is to increaseevaluative diversity (Stirling, 2008: 268), opening up the wickedness and the ambiguities of transition issues. 
@kemp2022DiversifyingDeepTransitions [113] -->

<!-- Do I need patents to actually show the entire funnel? -->

Recent scholarship has called for advancing research fronts instead of remaining entrenched in contested discourse positions [@allain2022BioeconomicsVsBioeconomy].
Here is how I propose to measure how the visions have provided directionality to the Swedish Transition to a bioeconomy.
Directionality as influencing the direction of search.
I then measure the effectivness of the directionality provided by different visions in actually producing innovation in the direction of search.
The emergence of visions is at least partially grounded in development paths taken by system actors.
These then serve as important vehicles to create legitimacy of the chosen paths and further provide directionality for the search.

@@

Based on the analyses of competing bioeconomy visions in @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy, I develop and apply a matrix of characterstics for each vision.
<!-- TODO: Matrix That matrix is presented in @tbl- -->
Reading the source texts of innovations identified from SWINNO, I classify an innovation as being *aligned* with a vision if it fulfills at least one criteria in the matrix.
This means that one innovation can be aligned with more than one vision at a time, for example, if it uses biological processes such as bacteria (biotechnology) to refine a wood based material to replace a fossil input elsewhere in the economy (bioresource vision).
Not all innovations fullfill at least one criteria, yet still originate or are used within the analytical boundaries of the forest based bioeconomy.
I classify these innovations as Vision Neutral.
Examples include innovations aimed at the health and saftey of employees 
<!-- TODO enter source -->
, or general improvements to the longevity of materials and tools used somewhere in the value chain of the forest bioeconomy.
The majority of innovations in this category are of the latter nature.

Innovations were classified as belonging to a bioeconomy vision if they fulfilled one, or more criteria outlined in @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy.
@tbl-bugge and @tbl-vivien show the summary of bioeconomy visions presented in the cited papers.

**DECISION MATRIX HERE**

For the remainder of the paper the nomenclature of @bugge2016WhatBioeconomyReview is adopted.
Vivien et al.'s [-@vivien2019HijackingBioeconomy] Type I bioeconomy is taken to correspond with the bioecology vision, Type II and Type III to correspond with the bioresource and biotechnology vision, respectively.
The analysis of bioeconomy discourse which revealed different visions, highlights that the visions have emerged at different times and with different actors at their centers [@vivien2019HijackingBioeconomy, @bugge2016WhatBioeconomyReview].
@vivien2019HijackingBioeconomy trace the introduction of the term bioeconomics to the 1920s and a work on fishery by the Russian biologist Baranoff.
The term bioeoconomics which they label as Type I bioeconomy, was introduced by Georgescu-Roegen, in [-@georgescu-roegen1975BioeconomicAspectsEntropy].
@vivien2019HijackingBioeconomy argue that the term biotechnology gradually emerged between 1990 and 2000, in the wake of enthusiasm about biotechnological advances, such as the discovery of DNA.
It gained popularity and international importance when the OECD began increasingly supporting and reshaping its meaning at the turn of the millennium [@vivien2019HijackingBioeconomy].
According to them, the last concept to enter the discourse was the bioresource oriented bioeconomy.
While biomass has been used throughout history in various ways, the renewed interest with the intention to replace fossil inputs, for example in European Commission communication since 2010.
However, @vivien2019HijackingBioeconomy point to the long history of biorefinery as a core aspect of this vision, which was discussed as early as the 1930s.
Hence, while the organization of the discourse into the labels biotechnology, bioecology and bioresource vision is a very recent research outcome, the underlying concepts have a richer history.
This alleviates concerns that using these labels retroactively confines the innovations to current policy terms.

@@

@tbl-vision-examples shows five examples of classified innovation.
This selection also highlights that multiple labels can be attached to one innovation.
The last innovation, describing a waterproof paper, is classified as a bioresource vision innovation because it describes a process by which biomass is converted and upgraded.
It is also classified as a biotechnology innovation since, according to the innovation source article, the process mimics trees' microfiber formation.
Lastly, there is also the possibility that an innovation cannot be placed into at least one of the visions.
This is the case for innovation which brings general technological improvements, purely advances economic performance, or relates to a general purpose technology such as appliance of computers.
These innovations are labeled as Vision Neutral, which does not imply that the innovation is not used, or originates in the forest based bioeconomy as defined above.
An illustrative example of such an innovation are improved work boots, which promote safety of forest workers but do not correspond with any of the criteria in @tbl-bugge or @tbl-vivien.

```{python}
# | label: tbl-vision-examples
# | tbl-cap: "List of Eco-Innovation Labels"

@@

This initial list was tested on a subset of innovation in SWINNO and augmented with 15 categories.
Because the list of eco-innovation types was relatively exhaustive from the beginning, few adjustments had to be made.
These include: expanding the labeling for other sustainability criteria, such as human health and safety, and specific forest related aspects / tasks to gain a more detailed understanding of an innovations aim.
For example, some innovations aim to protect harvested timber, either at harvest, during transport or storage.
While @tbl-full-codes shows all new added labels, in the following text those that do not clearly represent an eco-innovation -- following @garcia-granero2018EcoinnovationMeasurementReview "…fewer adverse effects on the environment and more efficient use of resources" (p. 305) -- are excluded.
These excluded labels are Gen-manipulated plant, Increased plant yield, Gen-manipulated animal (incl new breed), Human Health and Safety, Automatization, Precision, and Biocide.

@@

```{python}
# | label: fig-bioeconomy-share
# | fig-cap: "Bioeconomy Innovation Trends in SWINNO (Uncleaned)"
# | fig-subcap:
# |   - "Bioeconomy Innovation Count"
# |   - "Bioeconomy Innovation Share in %"
# | layout-ncol: 1

@@

bioeconomy_visions.loc[
    bioeconomy_visions["bioeconomy_vision"] == "Not Bioeconomy",
    ["bioeconomy_vision"],
] = "Vision Neutral"

@@

```{python}
# | label: fig-bioeconomy-trends
# | fig-cap: "Count of Bioeconomy Visions by Year"

@@

@fig-bioeconomy-trends illustrates that the peaks in innovation output, identified in @fig-bioeconomy-share and discussed above, were driven by innovation related to the Bioresource Vision as well as Vision Neutral.
While the Bioresource Visions did strongly contribute to the first peak, the second appears to have been primarily driven by innovation which cannot be associated as belonging to any vision of a bioeconomy.
This raises a couple of questions regarding the effectiveness of past and current policy.
It does add, for example, additional heft to the argument put forth by @holmgren2022WhoseTransformationThis that industry interest dominate the transformation to a bioeconomy.
Further, the persistent dominance of the bioresource vision over the full study period suggests that there might not be much effort, at least in terms of production of innovations, put into creating a transformation towards a bioeconomy.

```{python}
# | label: fig-vision-heatmap
# | fig-caption: "Sector's Contribution to Bioeconomy Visions"
# | include: false

@@

# sectors.loc[sectors["vision"] == "Not Bioeconomy", "vision"] = "Vision Neutral"
